---
title: "Importing, Modifying, Filtering"
author: "William Lizr"
date: "2023-01-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r cars}
summary(cars)
```

```{r pressure, echo=FALSE}
plot(pressure)
```

# My Notes Start Here:
## some useful basic commands:
```{r}
# a period sucks the 'piped' data into a certain place in the next command
mtcars %>% dplyr::filter(hp >= 50) %>% summarize(. , upper_mean = mean(hp))
# here the data is being sucked into the '.' place (it would do this automatically but sometimes this is important)

# to get all the unique variables in a factor, use unique()
# to select a certain column from a data frame, use select()
iris %>% select(Species) %>% unique()

#to get only certain rows, use slice()

cars <- mtcars
cars <- cars %>% slice(., c(2,3,4,5))
cars
#we will get 4 rows only, the ones selected above

# to read in an Excel file, use the readxl() package, for CSV use read.csv() or read_csv()
# for a base R dataframe we use read.csv()
# for a tibble we use read_csv()

# from readxl()
read_excel(	data , sheet = "" , range = cell_cols("X:Y") )

# exotic pipe operators
' %<>% ' = ' %>% ' + ' <- '
# use at the beginning of a line to input and rewrite a df at the end
# d_frame <- d_frame %>% = d_frame %<>%
cars <- mtcars
cars %<>% select(cyl) %>% unique()
cars

```

# data.frame vs Tibble vs data.table
```{r}
#to check class use the class() function
# for a base  R Dataframe we use read.csv()
# for a       Tibble we use read_csv()

# data frame (df) vs vector:
# df is a table essentially
# vector is an array essentially

#when using the group_by() command, we get a grouped_tibble
tibble(iris) %>% group_by(Species) %>% class()

#to convert to tibble, use as.tibble()
class(iris)
iris_tibble <- as_tibble(iris)
iris_tibble
class(iris_tibble)
# this tibble won't be grouped

```

# generating sample data
```{r}
# sample numerical integer data
v <- sample(x = 1:10, size = 1000, replace = TRUE)
hist(v)

#sample numerical data from a normal distribution
v <- rnorm(10000, mean = 100, sd = 25)
hist(v)
#to see all distributions: use ?distribution

# find 15 cars from defined dataset
lux_cars <- c(rep("Seat", 80), rep("Lambo", 10), rep("Skoda", 50)) %>% sample(. , size = 50, replace = TRUE) %>% as.vector()

# generate index values
index <- seq(1, length(lux_cars), by = 1)

# create a dataframe:
df <- data.frame(index, lux_cars) %>% as_tibble()
df

lux_cars
as_tibble(lux_cars)
colnames(lux_cars)

# to generate samples from a normal distribution, use "rnorm()"
gen_dat <- as_tibble(round(rnorm(n = 10000, mean = 100, sd = 15), digits = 2))
gen_dat

#getting counts of unique values (and retyping to a tibble once again)
gen_dat <- as_tibble(table(gen_dat))
gen_dat

# converting a whole factor to a numeric
transform(gen_dat, value = as.numeric(value))
```
# subsetting
```{r}
df <- iris %>% as_tibble()

#extract the first column like this:
df[[1]]
#or this:
df[,1]

```

# intro to sapply
```{r}

new_iris <- iris %>% as_tibble()

# getting the class for all columns in new_iris
sapply(new_iris, class)

# once again, the '.' pulls the data into the right spot after the next pipe
new_iris <- new_iris %>% .[[2]]

new_iris
```

# concatenating chars to a String
```{r}

paste0("Hi", "dear", "how", "are", "you", "today")

# using a separator
paste0(c("Hi", "dear", "how", "are", "you", "today"), collapse = "--")

```

# Using the mutate() function/verb
```{r}
# converting a column to a z-score
df <- iris %>% as_tibble() 

df <- df %>% mutate("z-score-sepal-length"=(Sepal.Length-mean(Sepal.Length))/sd(Sepal.Length))
df

```

# date (dd/mm/yyyy) as a data type
```{r}
#first, lets generate some data
allyears <- seq(1970, 2023, by = 1)
allmonths <- seq(1, 12, by = 1)
alldays <- seq(1,31, by = 1)

#define how many observations we need
pred_length <- 25

# here we create sample vectors, which pull their values from the generated dates and months
yr <- c(sample(allyears, size = pred_length, replace = TRUE))
mnth <- c(sample(allmonths, size = pred_length, replace = TRUE))
day <- c(sample(alldays, size = pred_length, replace = TRUE))

#we combine the vectors into a data frame, then a tibble
date_table <- data.frame(mnth, day, yr) %>% as_tibble()
date_table

#formatting as dates:

# we put together each of the day, month and year columns
concat_table <- date_table %>% mutate(. , "comb_date" = paste0(date_table$day,"-", date_table$mnth, "-" ,date_table$yr))
concat_table

# we put use the * lubridate * package to transform create a new column of data type 'date'
library(lubridate)
concat_table_dated <- concat_table %>% mutate(. , "dated_combo" = lubridate::dmy(concat_table$comb_date))
concat_table_dated


# to delete the redundant columns, lets use subsetting

#first, make an index vector
index_v <- seq(1, pred_length, by = 1)

cleaned_dates <- concat_table_dated %>% .[[5]] %>% data.frame(index_v, .) %>% as_tibble()
cleaned_dates

#change the column names to be nice and tidy
colnames(cleaned_dates) <- c("participant no.","date")
```

# Using the separate() function/verb
## under the dplyr package
```{r}
# separating data into separate columns when merged
# lets separate the data we previously made into dates from cleaned_dates

separated_dates <- cleaned_dates %>% separate(., "date", c("year", "month","day"), sep = "-")

#if the data isnt unifrom, we have to use the argument 'extra = merge', so that overflow data is put into the last column
```

# we can also use the dplyr library between deliminators like brackets with the extract() function
```{r}

```

# Using the filter() verb
```{r}
# to get rid of missing values 
df_clean <- mtcars %>% dplyr::filter(!is.na(mtcars$cyl))
df_clean

# lets import a base R dataset
mtcars_offl <- mtcars
# and make some values go missing...
mtcars_offl[2:6,3] <- NA

#to make sure each row has all values and doesn't contain an NA use the complete.cases() function
df_complete <- mtcars_offl %>% dplyr::filter(., complete.cases(mtcars_offl))
df_complete
```

# How to export to a csv
## part of the readr library
```{r}
#to export to a new file
df_clean %>% write_csv(., file = "cleaned_data.csv")

#to export to an existing file name, we can append the data
df_clean %>% write_csv(., file = "cleaned_data.csv", append = TRUE)
#now the data is twice as long!
```

# how to export .rdata files
## important if you need to save for example a grouped tibble, or other special types of data so that they retain their form
```{r}
# to save, the .rdata file name has to be the same as the variable name from the environment
car_data <- mtcars
car_data %>% save(file = "car_data.rdata")
```


# how to load .rdata files
## !doesnt work
```{r}

new_df <- load(file = "car_data.rdata")
```




