---
title: "Intro, geometry types, aesthetics"
author: "William Lizr"
date: "2023-01-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Tibble vs Tribble             ==============================================
## a tibble has column TYPES
## a tribble creates a tibble by inputting ROW data, not column data as vectors
```{r}


# creating a data frame
df1 <- data.frame(c("John", "David"), c("Lennon", "Copperfield"), c(85, 110))
# fixing the column names
colnames(df1) <- c("First","Last","IQ")



# making a tibble, fixing column names
df2 <- tibble(c("John", "David"), c("Lennon", "Copperfield"), c(85, 110))
colnames(df2) <- c("First","Last","IQ")
# fixing the type of of the third column dbl —> int
df2$IQ <- as.integer(df2$IQ)



# making a tibble by row (tribble command)
df3 <- tribble( ~First, ~Last, ~iQ, 
                "Mark", "Hammond", 150, 
                "Charlie", "Daniels", 129, 
                "Ramsey", "Christopher", 102)

df3$iQ <- as.integer(df3$iQ)
```

# always import ggplot2 ======================================================
```{r}
# and import the conflicted package to manage your conflicts and function collisions
library(conflicted)

# tidyverse imports ggplot2, and other useful operators like pipes
library(tidyverse)

# we are also going to use ggmap
library(ggmap)

```

# reading data with read_csv()
##     because of the underscore    -> creates a tibble
##     a dot                        -> creates a data.frame

```{r}
# import csv
college_df <- read_csv(file = "college_dataset.csv")
# we see it is a tibble
class(college_df)


# now we are going to factorize some columns
college_df <- college_df %>% mutate(region = as.factor(region),
                                    state = as.factor(state), control = as.factor(control), gender = as.factor(gender), highest_degree = as.factor(highest_degree))

# now we can see them as factors when we run summary()
summary(college_df)


# to remove strings from numericals, we FORCE them to be NUMERICAL types, then we remove NA values, which R generates to subsitute for values we can't convert
college_df <- college_df %>% mutate(loan_default_rate = as.numeric(loan_default_rate))
# we get a message:
# ' Warning: NAs introduced by coercion '

# now when we run summary we get the loan_default_rate as a numerical variable
summary(college_df)


# here we're going to get rid of the NA values in this column
college_df <- college_df %>% dplyr::filter(!is.na(loan_default_rate))
# we can also use the complete.cases 
college_df <- college_df %>% dplyr::filter(complete.cases(.))

# we also gon convert this bish to percent
college_df <- college_df %>% mutate(loan_default_rate = round(loan_default_rate*100, digits = 1))

summary(college_df)

```

# the grammar of graphics ====================================================
## data         the raw data
## geometries   shapes we use to visualize
## aesthetics   properties of geometries
## scales       mappings between geometries and aesthetics

```{r}
# 1: data
# to initialize an empty plot canvas, we use the ggplot() command
# even if we give it data, it doesn't know what to do with it
ggplot(data = college_df)

# 2: geometry
# then we add elements like geom_point(), geom_smooth(), geom_violin(), etc.
ggplot(data = college_df) + geom_point(mapping = aes(x = tuition, y = sat_avg))


# 3: we add some x and y axis labels and a chart title
ggplot(data = college_df) + geom_point(mapping = aes(x = tuition, y = sat_avg)) + labs(title = "correlation of tuition and average sat") + xlab("Tuition") + ylab("SAT score")

```


# Scatterplots
## using geom_point()
## basic properties: x = __ ; y = __ ;
### shape       shape of points by categorical variable
### color       point color
### size        point size
### alpha       transparency

# see above code block for initial example
```{r}

# important to note:
# the 'mapping = aes(...)' is where you put aesthetics or features of the GRAPH that should REFLECT DIFFERENCES IN THE DATASET
# if we just want to adjust OVERALL shape, color, alpha or size — then we have to put it outside the mapping

# outside
ggplot(data = college_df) + geom_point(mapping = aes(x = tuition, y = sat_avg), color = 'blue') + labs(title = "correlation of tuition and average sat") + xlab("Tuition") + ylab("SAT score")

# inside (linked to a categorical variable; here, we used a 4 layer categorical variable called region)
ggplot(data = college_df) + geom_point(mapping = aes(x = tuition, y = sat_avg, color = region)) + labs(title = "correlation of tuition and average sat") + xlab("Tuition") + ylab("SAT score")

```

# More geometries — LINES
## still specify X and Y axes for the lines to display

## geom_line() connects the dots
```{r}

ggplot(data = college_df, mapping = aes(x = tuition, y = sat_avg, color = region)) + geom_point() + geom_line()


```


## geom_smooth() fits a line to the points like a trend
```{r}

ggplot(data = college_df, mapping = aes(x = tuition, y = sat_avg, color = region)) + geom_point() + geom_smooth()

# to get rid of the standard error (SE) area around the 'smoothers', we use 'SE = FALSE'
ggplot(data = college_df, mapping = aes(x = tuition, y = sat_avg, color = region)) + geom_point() + geom_smooth(se = FALSE)

# lets also decrease the alpha of the points to be more readable
ggplot(data = college_df, mapping = aes(x = tuition, y = sat_avg, color = region)) + geom_point(alpha = 0.25) + geom_smooth(se = FALSE) 

```


# BARS AND COLUMNS
## geom_bar()
### specify X VALUE
### uses count() as the Y VALUE 

## geom_col()
### specify X VALUE
### allows user to specify the Y VALUE


# geom_bar()
```{r}



```


# geom_col()
```{r}


# custom Y VALUE
```


# HISTOGRAMS: using geom_histogram()
## continuous variable on the x axis is necessary, bins are a feature of that
```{r}

ggplot(data = college_df) + geom_histogram(mapping = aes(x = sat_avg))

# we can change the range of each 'block'(bin), to get a more granular graph:
ggplot(data = college_df) + geom_histogram(mapping = aes(x = sat_avg), binwidth = 3)
# or we can code the NUMBER OF BINS, instead of BIN WIDTH
ggplot(data = college_df) + geom_histogram(mapping = aes(x = sat_avg), bins = 30)
```

# JITTER PLOTS
## with geom_jitter()
```{r}
# first we need to see a bad use case of an alternative
# for example: geom_point()
ggplot(data = college_df) + geom_point(mapping = aes(x = control, y = tuition))
# this doesnt allow us to see the resolution of 

# here we have some more width to the data
ggplot(data = college_df) + geom_jitter(mapping = aes(x = control, y = tuition))


```

# BOXPLOTS
## with geom_boxplot()
```{r}
ggplot(college_df) + geom_boxplot(aes(x = control, y = tuition))

# we can also tweak the outlier presence
ggplot(college_df) + geom_boxplot(aes(x = control, y = tuition, fill = control), outlier.color = 'red')

# we can modify their shape, size, color, fill, etc


```












































































